On the Bit Complexity of Finding Points in Connected Components of a Smooth Real Hypersurface

### 2020-02-07 main.tex
Update main.tex
Updated Section 1 and Theorem 1.1

\paragraph*{Data structures.} As in references such
as~\cite{GiHeMoPa95,GiHaHeMoMoPa97,GiHeMoMoPa98,BaGiHeMb97,EMP}, the
input polynomial $f$ is given by means of a {\em straight-line
  program}, that is, a sequence of operations $+,-,\times$ that takes
as input $X_1,\dots,X_n$ and evaluates $f$; the {\em length} of a
straight-line program is simply the number of operations it performs.

Any polynomial of degree $d$ in $n$ variables can be computed by a
straight-line program of length $O\big ({{d+n} \choose n}\big)$:
evaluate all monomials of degree up to $d$ in $n$ variables, multiply
them by their respective coefficients and sum the result. However,
some inputs may be given by shorter straight-line program, and the
algorithm is able to benefit from this.

The output of the algorithm is a finite set in $\overline{\Q}{}^n$. To
represent it, we rely on a widely used data structure based on
univariate
polynomials~\cite{Kronecker82,Macaulay16,GiMo89,GiHeMoPa95,ABRW,GiHaHeMoMoPa97,GiHeMoMoPa98,Rouillier99}.
Consider a zero-dimensional algebraic set $S \subset \C^n$ defined
over $\Q$. A {\em zero-dimensional parametrization}
$\scrQ=((q,v_1,\dots,v_n),\lambda)$ of $S$ consists in polynomials
$(q,v_1,\dots,v_n)$, such that $q\in \Q[T]$ is monic and squarefree,
all $v_i$'s are in $\Q[T]$ and satisfy $\deg(v_i) < \deg(q)$, and in a
$\Q$-linear form $\lambda$ in variables $X_1,\dots,X_n$, such that
\begin{itemize}
\item $\lambda(v_1,\dots,v_n)=T q' \bmod q$;
\item we have the equality
  $S=\left \{\left(
      \frac{v_1(\tau)}{q'(\tau)},\dots,\frac{v_n(\tau)}{q'(\tau)}\right
    ) \ \mid \ q(\tau)=0 \right \}.$
\end{itemize}
The constraint on $\lambda$ says that the roots of $q$ are the values
taken by $\lambda$ on $S$. The parametrization of the coordinates by
rational functions having $q'$ as a denominator goes back
to~\cite{Kronecker82,Macaulay16}: as pointed out in~\cite{ABRW}, it
allows one to control precisely the size of the coefficients of
$v_1,\dots,v_n$.

@@

\begin{theorem}\label{theo:main}
  Suppose that $f\in\ZZ[X_1\hdots,X_n]$ is squarefree, satisfies
  $\deg(f) \leq d, \htt(f) \leq b$, and that $V(f) \subset \C^n$ is
  smooth.

  There exists a randomized algorithm that takes $f$
  as input and produces a zero-dimensional parameterization of a set
  that contains at least one point in each connected component of
  $V(f) \cap \R^n$.

  As part of the randomization, the algorithm chooses sets 
  \[
  S := \{1,2,\hdots,3\epsilon^{-1}6n^3(2d)^{2n}\}
  \]
  and
  \[
  T := \{1,2,\hdots,3\epsilon^{-1}nd^{2n}\}
  \]
  from which to sample from.

  With probability at least $1-\epsilon$, the output is correct. Otherwise, 
  the algorithm either returns a proper subset of the points, or FAIL.  In
  any case, the algorithm uses
  \[
O^{\sim}(d^{3n+1}(\log1/\epsilon)(b + \log1/\epsilon))
  \]
  bit operations. The polynomials in
  the output have degree at most $d^n,$ and height 
  \begin{align*}
  &O^{\sim}(d^{n+1}(b + \log 1 / \epsilon)). 
  \end{align*}
\end{theorem}

Here we assume that $f$ is given as a dense polynomial. However, it 
would be possible to take into account the runtime if $f$ was given 
by a straight-line program $\Gamma,$ where the length of $\Gamma$ would
be bounded by $d^n$ and would have integer constants of height at most
$b$. Here, $\Gamma$ could be computed in the trivial way, by 
computing the terms and summing them. 
The algorithm itself is rather simple. To describe it, we need to
define {\em polar varieties}, which will play a crucial role in this
paper. Let $V=V(f)$, for $f$ as in the theorem. For $i \in
\{1,\hdots,n-1\},$ denote by $\pi_i:\C^n \rightarrow \C^i$ the
projection $(x_1,\hdots,x_n) \mapsto (x_1,\hdots,x_i)$.  The $i$-th
\textit{polar variety} \[W(\pi_i,V) := \{\xb \in V~|~\dim \pi_i(T_\xb
V) < i\}\] is the set of critical points of $\pi_i$ on $V$.  It is
thus defined by the vanishing of \[f,\frac{\pa f}{\pa
  X_{i+1}},\hdots,\frac{\pa f}{\pa X_n}.\]

In general, we cannot say much about the geometry of
$W(\pi_i,V)$, but if we apply a generic change of coordinates $\mA$,
$W(\pi_i,V)$ is then known to be equidimensional of dimension $(i-1)$
or empty~\cite{BaGiHeMb97}, and to be in so-called {\em Noether
  position}~\cite{EMP} (background notions in algebraic geometry are
in~\cite{Mumford76,Shafarevich77,ECA}; we will recall key
definitions). If this is case, it suffices to choose arbitrary
$\alpha_1,\dots,\alpha_{n-1}$ in $\Q$, and solve the systems defined
by
\begin{equation}\label{eq:syst}
X_1-\alpha_1,\dots,X_{i-1}-\alpha_{i-1}, f, \frac{\pa f}{\pa
  X_{i+1}} ,\dots,\frac{\pa f}{\pa X_n},  
\end{equation}
for $i=1,\dots,n$.  They all admit finitely many solutions, and
Theorem~2 in~\cite{EMP} proves that the union of their solution sets
contains one point on each connected component of $V \cap \R^n$.

>>>>

\paragraph*{Data structures.} 
The output of the algorithm is a finite set in $\overline{\Q}{}^n$. To
represent it, we rely on a widely used data structure based on
univariate
polynomials~\cite{Kronecker82,Macaulay16,GiMo89,GiHeMoPa95,ABRW,GiHaHeMoMoPa97,GiHeMoMoPa98,Rouillier99}.
Consider a zero-dimensional algebraic set $S \subset \C^n$ defined
over $\Q$. A {\em zero-dimensional parametrization}
$\scrQ=((q,v_1,\dots,v_n),\lambda)$ of $S$ consists in polynomials
$(q,v_1,\dots,v_n)$, such that $q\in \Q[T]$ is monic and squarefree,
all $v_i$'s are in $\Q[T]$ and satisfy $\deg(v_i) < \deg(q)$, and in a
$\Q$-linear form $\lambda$ in variables $X_1,\dots,X_n$, such that
\begin{itemize}
\item $\lambda(v_1,\dots,v_n)=T q' \bmod q$;
\item we have the equality
  $S=\left \{\left(
      \frac{v_1(\tau)}{q'(\tau)},\dots,\frac{v_n(\tau)}{q'(\tau)}\right
    ) \ \mid \ q(\tau)=0 \right \}.$
\end{itemize}
The constraint on $\lambda$ says that the roots of $q$ are the values
taken by $\lambda$ on $S$. The parametrization of the coordinates by
rational functions having $q'$ as a denominator goes back
to~\cite{Kronecker82,Macaulay16}: as pointed out in~\cite{ABRW}, it
allows one to control precisely the size of the coefficients of
$v_1,\dots,v_n$.

@@

\begin{theorem}\label{theo:main}
  Suppose that $f\in\ZZ[X_1\hdots,X_n]$ is squarefree, satisfies
  $\deg(f) \leq d, \htt(f) \leq b$, and that $V(f) \subset \C^n$ is
  smooth. Also suppose that $0 < \epsilon < 1.$

  There exists a randomized algorithm that takes $f$ and $\epsilon$
  as input and produces $n$ zero-dimensional parameterizations whose reunion includes at least one point in each connected component of
  $V(f) \cap \R^n$, with probability at least $1-\epsilon$. Otherwise, 
  the algorithm either returns a proper subset of the points, or FAIL.  In
  any case, the algorithm uses
  \[
O^{\sim}(d^{3n+1}(\log1/\epsilon)(b + \log1/\epsilon))
  \]
  bit operations. The polynomials in
  the output have degree at most $d^n,$ and height 
  \begin{align*}
  &O^{\sim}(d^{n+1}(b + \log 1 / \epsilon)). 
  \end{align*}
\end{theorem}

Here we assume that $f$ is given as a dense polynomial. However, it 
would be possible to take into account the runtime if $f$ was given 
by a {\em straight-line
  program}, that is, a sequence of operations $+,-,\times$ that takes
as input $X_1,\dots,X_n$ and evaluates $f$; the {\em length} of a
straight-line program is simply the number of operations it performs. Straigh-line programs are described in references such
as~\cite{GiHeMoPa95,GiHaHeMoMoPa97,GiHeMoMoPa98,BaGiHeMb97,EMP}. The length of the straight-line program for $f$ would
be bounded by $d^n$ and would have integer constants of height at most
$b$. Any polynomial of degree $d$ in $n$ variables can be computed by a
straight-line program of length $O\big ({{d+n} \choose n}\big)$:
evaluate all monomials of degree up to $d$ in $n$ variables, multiply
them by their respective coefficients and sum the result. However,
some inputs may be given by shorter straight-line program, and the
algorithm is able to benefit from this.

The algorithm itself is rather simple. To describe it, we need to
define {\em polar varieties}, which will play a crucial role in this
paper. Let $V=V(f)$, for $f$ as in the theorem. For $i \in
\{1,\hdots,n-1\},$ denote by $\pi_i:\C^n \rightarrow \C^i$ the
projection $(x_1,\hdots,x_n) \mapsto (x_1,\hdots,x_i)$.  The $i$-th
\textit{polar variety} \[W(\pi_i,V) := \{\xb \in V~|~\dim \pi_i(T_\xb
V) < i\}\] is the set of critical points of $\pi_i$ on $V$.  It is
thus defined by the vanishing of \[f,\frac{\pa f}{\pa
  X_{i+1}},\hdots,\frac{\pa f}{\pa X_n}.\]

In general, we cannot say much about the geometry of
$W(\pi_i,V)$, but if we apply a generic change of coordinates $\mA$,
$W(\pi_i,V)$ is then known to be equidimensional of dimension $(i-1)$
or empty~\cite{BaGiHeMb97}, and to be in so-called {\em Noether
  position}~\cite{EMP} (background notions in algebraic geometry are
in~\cite{Mumford76,Shafarevich77,ECA}; we will recall key
definitions). If this is case, it suffices to choose arbitrary
$\alpha_1,\dots,\alpha_{n-1}$ in $\Q$, and solve the systems defined
by
\begin{equation}\label{eq:syst}
X_1-\sigma_1,\dots,X_{i-1}-\sigma_{i-1}, f, \frac{\pa f}{\pa
  X_{i+1}} ,\dots,\frac{\pa f}{\pa X_n},  
\end{equation}
for $i=1,\dots,n$.  They all admit finitely many solutions, and
Theorem~2 in~\cite{EMP} proves that the union of their solution sets
contains one point on each connected component of $V \cap \R^n$.
